---
title: "Data 607 Week 3 Assignment"
author: "Judd Anderman"
date: "September 18, 2016"
output: html_document
---

#### _Automated Data Collection with R_, Ch. 8: Regular expressions and essential string functions 
This week's assignment applies the text processing capabilities of regular expressions and the string manipulation functionality offered by Hadley Wickham's __stringr__ package to problems 3, 4, and 9 from Chapter 8 in _Automated Data Collection with R_ (2015).

<br>

#### Load required package
```{r load-libs}
library(stringr)
```

<br>

#### Problem 3
3\. Copy the introductory example.  The vector `name` stores the extracted names.

```{r exercise-3}
raw.data <- "555-1239Moe Szyslak(636) 555-0113Burns, C. Montgomery555-6542Rev. Timothy Lovejoy555 8904Ned Flanders636-555-3226Simpson, Homer5553642Dr. Julius Hibbert"   

name <- unlist(str_extract_all(raw.data, "[[:alpha:]., ]{2,}"))
phone <- unlist(str_extract_all(raw.data, "\\(?(\\d{3})?\\)?(-| )?\\d{3}(-| )?\\d{4}"))

name
```

3.1. Use the tools of this chapter to rearrange the vector so that all elements conform to the standard `first_name` `last_name`.

```{r ex-3.1}
splitname <- str_split(name, ", ", simplify = TRUE)
first_last <- str_c(splitname[, 2], splitname[, 1])
first_last <- str_replace(first_last, "([a-z])([A-Z])", "\\1 \\2")
first_last

no_title <- str_replace(first_last, "[[:alpha:]]{2,}\\. ", "")
no_title 

first_name <- str_extract(no_title, "^[[:alpha:]]+\\.?")
last_name <- str_extract(first_last, "[[:alpha:]]+$")

no_mid_name <- str_c(first_name, last_name, sep = " ")
no_mid_name
```

3.2. Construct a logical vector indicating whether a character has a title (i.e., `Rev.` and `Dr.`).

```{r ex-3.2}
title_test <- str_detect(first_last, "[[:alpha:]]{2,}\\.")
title_test

title <- str_extract(first_last, "[[:alpha:]]{2,}\\.")
title
```

3.3. Construct a logical vector indicating whether a character has a second name.

Assuming "second name" refers to surname:
```{r ex-3.3.1}
last_test <- str_detect(first_last, "[[:alpha:]]+$")
last_test

last_name
```

Assuming "second name" refers to middle name:
```{r ex-3.3.2}
second_test <- str_detect(first_last, "(?<!([[:alpha:]]{2,3}\\.)) [[:alpha:]]+ ")
second_test

second_name <- str_trim(str_extract(first_last, "(?<!([[:alpha:]]{2,3}\\.)) [[:alpha:]]+ "))
second_name
```
  
Creating a _Simpsons_ characters' names and phone numbers data frame:
```{r simpsons-df}
pnum <- "\\(?([0-9]{3})?\\)?[- ]?([0-9]{3})[- ]?([0-9]{4})"

names_df <- cbind(last_name, first_name, middle_name = second_name, title, area_code = str_match(phone, pnum)[, 2], phone = str_c(str_match(phone, pnum)[, 3], "-", str_match(phone, pnum)[, 4]))
                  
knitr::kable(names_df)
```

<br>

#### Problem 4
4\. Describe the types of strings that conform to the following regular expressions and construct an example that is matched by the regular expression.

4.1. `[0-9]+\\$`

This regular expression describes a string of one or more digits followed by a dollar sign, for example, `958$`.

4.2. `\\b[a-z]{1,4}\\b`

This regular expression describes a word of one to four consecutive lowercase alphabetic characters, in other words, a string of one to four lowercase letters with word boundaries on both sides.  An example is `pear`.

4.3. `.*?\\.txt$`

This regular expression describes a string of any number of characters ending in ".txt", i.e., a file or path name with a .txt extension.  For example, `folder/file.txt`.  

4.4. `\\d{2}/\\d{2}/\\d{4}`

This regular expression describes 2 digits followed by a forward slash, 2 more digits, another forward slash, and four more digits, in other words, a date in the format `12/21/1994` or `18/09/2016`.

4.5. `<(.+?)>.+?</\\1>`

This regular expression describes an xml or html element with start and end tags, without attributes, and with variable content of length of at least one character.  An example is `<data> 123 </data>`.

In the following code, the examples are checked to verify that they conform to the provided regular expressions:
```{r regex-examples}
examples <- c("958$", "pear", "folder/file.txt", "12/21/1994", "18/09/2016", "<data> 123 </data>")

#4.1
str_extract(examples, "[0-9]+\\$")
#4.2
str_extract(examples, "\\b[a-z]{1,4}\\b")
#4.3
str_extract(examples, ".*?\\.txt$")
#4.4
str_extract(examples, "\\d{2}/\\d{2}/\\d{4}")
#4.5
str_extract(examples, "<(.+?)>.+?</\\1>")
```

<br>

#### Problem 9
9\. The following code hides a secret message.  Crack it with `R` and regular expressions.  _Hint: Some of the characters are more revealing than others!  The code snippet is also available in the materials at [www.r-datacollection.com](http://www.r-datacollection.com)_.

```{r exercise-9}
hidden <- "clcopCow1zmstc0d87wnkig7OvdicpNuggvhryn92Gjuwczi8hqrfpRxs5Aj5dwpn0TanwoUwisdij7Lj8kpf03AT5Idr3coc0bt7yczjatOaootj55t3Nj3ne6c4Sfek.r1w1YwwojigOd6vrfUrbz2.2bkAnbhzgv4R9i05zEcrop.wAgnb.SqoU65fPa1otfb7wEm24k6t3sR9zqe5fy89n6Nd5t9kc4fE905gmc4Rgxo5nhDk!gr"

unlist(str_extract_all(hidden, "[A-Z]"))
unlist(str_extract_all(hidden, "[a-z]"))
unlist(str_extract_all(hidden, "[0-9]"))
```

The hidden message is revealed by extracting just the uppercase characters from the given string as below:
```{r code-reveal}
code <- str_c(unlist(str_extract_all(hidden, "[A-Z]")), collapse = "")
code
```